---
title: "README"
output: 
  md_document:
    toc: yes
---


```{r setup, include=FALSE, cache = TRUE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, include=FALSE}
library(dplyr)
library(ciTools)
library(MASS)
library(tidyverse) 
library(ggplot2)
library(prettyGraphs)
library(ggfortify)
library(forecast)
library(lmtest)
```



# Global COVID-19 Response
_Last updated: 26 February 2021_

## Table of Contents
- [About](#About)
- [Goals](#Goals)
- [Modeling technique](#Modeling-technique)
  - [Facility-level models](#Facility-level-models)
  - [Aggregated models](#Aggregated-models)
  - [Missing data considerations](#Missing-data-considerations)
- [Overview of folders and files](#Overview-of-folders-and-files)
- [Examples](#Examples)

## About:
This repository contains code to follow the Data Processing Pipeline for the Global Covid-19 Syndromic Surveillance Team - a partnership between sites at Partners in Health, the Global Health Research Core at Harvard Medical School, and Brigham and Women's Hospital. The data has been modified to respect the privacy of our sites, in hopes that other groups can benefit from the functions we have written.

This repository contains data, code, and other items needed to reproduce this work. Further explanation of outputs and their construction is given in the "Overview of folders and files" section, which includes detailed explanations of the functions we have written.


## Goals:
The main goal of the Global COVID-19 Syndromic Survillance Team is to monitor changes in indicators that may signal changes in COVID-19 case numbers in health systems from our seven partnering countries: Haiti, Lesotho, Liberia, Malawi, Mexico, and Rwanda. This is accomplished through establishing a baseline using prior data, and monitoring for deviations for relevant indicators. The data visualization tools created using our functions allow identification of local areas that are experiencing upticks in COVID-19-related symptoms.

## Modeling technique:
The process starting with the raw data and finishing with the various outputs is referred to as the Data Processing Pipeline (see Figure 1 below):

After data has been cleaned, it is processed according to the level it is available at (either on a facility of county/district basis) for each indicator. This is done by taking data from a historic baseline period, and then projecting it into the evaluation period. This then is compared to the observed counts/proportions. For our purposes, we have defined the baseline period to be data from January 2016-December 2019. 

The functions included in this repository focus on the modeling and processing stages.

### Facility-level models:

For facility-level assessments, we fit a generalized linear model with negative binomial distribution and log-link to estimate expected monthly counts. Only data from the baseline period will be used to estimate the expected counts:

$$ \log(E[Y | year, t ]) = \beta_0 + \beta_1year + \sum_{k=1}^{3} \beta_{k1} cos(2 \pi kt/12) + \beta_{k2} sin(2 \pi kt/12) $$
where Y indicates monthly indicator count, t indicates the cumulative month number. The year term captures trend, and the harmonic term captures seasonality. This model is an adaptation of that proposed by Dan Weinberger lab (https://weinbergerlab.github.io/ExcessILI/articles/PImortality.html). If data is available on a more granular level, then weekly or daily terms could be added to the equation to capture other types of trend. To calculate the prediction intervals, we used ciTools R package (https://cran.r-project.org/web/packages/ciTools/ciTools.pdf).
 
For proportions, in which the numerator is indicator counts and the denominator is outpatient visits, we produced similar prediction intervals using the following procedure: we performed a parametric bootstrap procedure that generates random monthly indicator counts from the prediction intervals described above and kept the total outpatient visits fixed. This gives empirical estimates and prediction intervals for proportions. If there were missing values for the monthly outpatient visit count, instead of deleting those months and doing a complete-case analysis which would waste existing indicator count data, we performed an imputation procedure as follows: first, we fit the aforementioned model for outpatient visits instead of indicator counts, and using that model’s estimates, imputed the missing denominator values. Then, we can do the parametric bootstrap procedure with the additional step of randomly imputing missing denominator values in order to account for variation and uncertainty in these imputed outpatient values. 


### Aggregated models:

It was also of interest to perform syndromic surveillance at aggregated levels (e.g. district or county). If there was no missing monthly data, one could simply sum the indicator counts across all facilities within a district (or county) and fit the above model. However, data often contains months with missing counts at the facility-level. We used a parametric bootstrap to impute the missing values from the facility-level models in the previous section. We drew realizations of the ARI counts for each month and each facility and then summed these values for an aggregated estimate. We repeated this procedure 500 times and took the 2.5th and 97.5th percentiles to create 95% prediction intervals. For region-level proportions, the number of outpatient visits can be summed across facilities and a proportion can be computed. If there are missing values in the outpatient visits, another step can be included in the above parametric bootstrap procedure where missing outpatient visits are generated from fitting the above model and where Y indicates monthly outpatient visit count.


### Deviations and data visualizations: 
We defined a deviation as the difference between the predicted and observed count. To allow interpretation across facilities and regions of different sizes, we divided by the predicted count for a scaled deviation measure ((expected-observed)/expected), where positive values mean that the observed number of acute respiratory infections is higher than expected, potentially indicating circulation of COVID-19. In our data visualizations, we report this scaled deviation measure in addition to indicating if the observed count falls outside of the 95% prediction interval. We provide 2 ways here to visualize the results: time series plots and tiled heatmaps, with examples shown below.

### Missing data considerations:
We excluded facilities from our analysis for two reasons: (1) missing dates in the baseline period (creation of the expected counts model) (2) missing observed counts in the evaluation period.

For the first reason, facilities with high levels of missing data (more than 20% of baseline dates missing) were excluded. Although there are statistical methods that can handle missing time series data, we decided to only include sites that demonstrated ability to collect data over time. A complete case (time) analysis was conducted with included facilities, which assumes that the counts were missing completely at random (MCAR). Specifically, we assumed the reason for missing counts was independent of time and count value. If the MCAR assumption was violated and we had information about the missing data mechanism, one could impute values for the missing data and report unbiased expected counts and correct inference.

For the second reason, facilities with ANY missing monthly data during the evaluation period (January 2020 onward) were removed. As the syndromic surveillance exercise hinges on comparing the observed counts to the expected and flagging for deviations, we require complete observed data during this period. In this context, it would be invalid to impute observed counts based on information from the baseline period. In theory, one could attempt to impute the observed count based on information during the evaluation period.

## Overview of folders and files:
### Data
This folder contains example data used to demonstrate functions.
#### data.example_singlecounty.rds
The facility-level dataset used to demonstrate the functions throughout this repository. Note- specific names and numbers have been altered to respect the privacy of our sites.

### R
This folder contains the functions used to create the key data visualization figures and maps.


### Figures
This folder contains figures that have been included in README.md.


## Examples

#### Loading Data and Functions

```{r,warning=FALSE,message=FALSE}
source("R/model_functions.R")
source("R/model_figures.R")
source("R/model_diagnostics.R")
```


```{r}
data <- readRDS("data/data_example.rds")
```


```{r}
head(data)
```

The example data contains two health facilities: "Facility K" and "Facility Q". The indicator of interest is acute respiratory infections (first column: ari_cases) with total outpatient visits (total_visits)--a measure of healthcare utilization. 

#### Example 1: Single Facility

We take an example facility--"Facility K", run the facility-specific model for one specific indicator--"ari_cases", with the associated denominator--"total_visits" and other parameter specifications as seen below, and look at the results through the counts and proportion lenses. 

```{r,warning=FALSE,message=FALSE}
# Declare this for all functions
extrapolation_date <- "2020-01-01"

# Run Facility-level Model
fit.site.specific.denom.pi(data=data,
                           site_var="facility",
                           site_name="Facility K",
                           extrapolation_date=extrapolation_date,
                           indicator_var="ari_cases",
                           denom_var="total_visits", 
                           date_var="date",
                           R=500) -> single.facility.results
```

The dataframe *example_1_results* contains the following fields: observed counts, estimated counts, prediction intervals for counts, observed proportion, estimated proportion, and prediction intervals for proportions.


```{r,warning=FALSE,message=FALSE}
head(single.facility.results)
```

We can view plots of the results for Facility K, for counts and for proportions, by running the following function and passing in the facility-specific model results dataframe as the first parameter and either "count" or "proportion" as the second parameter. 


##### Single Facility Counts Results

```{r}
plot_site(single.facility.results,type="count", title="Acute Respiratory Infections at Facility K")
```


The observed count is given by the __black line__ (raw data from DHIS2). The expected (predicted) count is given by the <font color='red'>__red line__</font> with 95% prediction intervals in light red (using the model described above).


##### Single Facility Proportions Results

```{r}
plot_site(single.facility.results,type="proportion", title="Acute Respiratory Infections at Facility K")
```



#### Example 2: Fit multiple facilities at once 

If there are multiple facilities, we may wish to fit multiple models at one time. Here, we provide code to apply over the function. Note that the same logic can be applied for multiple indicators across multiple sites. 

```{r,cache=TRUE}
# get all facilities
all_sites <- data %>% distinct(facility) %>% pull()

# loop over all  facilities
do.call(rbind, lapply(all_sites,function(x){
      fit.site.specific.denom.pi(data=data,
                              site_name=x,
                              extrapolation_date=extrapolation_date,
                              indicator_var="ari_cases",
                              denom_var="total_visits",   # corresponding denominator indicator needed for proportions
                              site_var="facility",
                              date_var="date",
                              R=500)
  })
) -> facility.results
```

The dataframe *facility.list* combines the results for each facility. 


```{r,cache=TRUE}
head(facility.results)
```

We can view plots of the results for both facilities together by running the following function and passing in the *facility.results*  dataframe as the first parameter and either "count" or "proportion" as the second parameter. 

```{r}
plot_facet(facility.results,type="count")
```

Also useful are heatmaps, especially for comparison between the different facilities using the deviation metric as described above. Again, the first parameter in the function below is the model output, and the second parameter is either "count" or "proportion":

```{r}
plot_heatmap(facility.results,type="count")
```

__Note:__ the black border boxes indicate the count (or proportion) falls outside the prediction interval.

#### Example 3: Aggregated results

We now wish to calculate a total count for the both facilities. Note that we decide not to sum the ARI cases and total visit counts in the two facilities because there are several values missing, which would result in an underestimated count during these months if ignored (i.e. assumed to be zero). 

```{r}
data %>% filter(is.na(ari_cases) | is.na(total_visits))
```


```{r,cache=TRUE}
fit.aggregate.pi.boot(data,
                      indicator_var = "ari_cases",
                      denom_var = "total_visits",
                      date_var = "date",
                      site_var = "facility",
                      R=500) -> aggregate.results
```

The fields in the output are the same as for a facility-level model.

```{r,cache=TRUE}
head(aggregate.results)
```

We can view similar plots as for the facility-specific plots above but now for the county of interest, using the same plotting function:

```{r}
plot_site(aggregate.results, "count", title="Facility K and Q Aggregated Results")
```

#### Example 4: Model diagnostics

We recommend reading (https://otexts.com/fpp2/residuals.html) for an overview on evaluating time series models. We provide code to replicate these plots and tests using the output from our main functions. We show at the residual plots using the output from our main functions. We do this for the aggregate results and then for the facility-level results for County K. 

```{r}
plot_residuals(aggregate.results,
               type="count",
               extrapolation_date="2020-01-01",
               title="Residuals from Aggregated model")
```

```{r,warning=FALSE}
plot_residuals(single.facility.results,
               type="count",
               extrapolation_date="2020-01-01",
               title="Residuals from County K model")
```

We then look at the Partial Autocorrelation Function and Autocorrelation Function plots for County K. The same code can be used for the aggregated results. 

```{r,warning=FALSE}
plot_pacf(single.facility.results,
          type="count",
          extrapolation_date="2020-01-01",
          title="PACF for County K")
```

```{r,warning=FALSE}
plot_acf(single.facility.results,
         type="count",
         extrapolation_date="2020-01-01",
         title="ACF for County K")
```

Lastly, we can conduct the Bresuch-Godfrey Test for the facility-level models (order = 12 for monthly data). This code will apply over multiple site (or faiclity) results. This provides a way to rapidly identify problematic model fits for further investigation.

```{r,warning=FALSE}
bgtest_multiple_pval(data=data,
                     indicator_var = "ari_cases",
                     denom_var = "total_visits",
                     date_var = "date",
                     site_var = "facility",
                     extrapolation_date = "2020-01-01",
                     type = "count") 
```

A p-value less than 0.05 indicates evidence of residual correlation. As such, we should further investigate the model fit for County Q. There is no evidence of residual correlation in Facility K. 